Tutorial: Quantifying Spatially Resolved Hydration Thermodynamics Using Grid Inhomogeneous Solvation Theory [Article v\versionnumber]

### 2024-06-27 manuscript/manuscript.tex
Removed unnecessary paragraph formating

On the other hand, methods based on IST generally use a molecular solvent distribution obtained from an MD simulation to compute the free energy of hydration from its enthalpic and entropic contributions, providing easier interpretability of the resulting distributions.
Examples include grid inhomogeneous solvation theory (GIST) \cite{Nguyen2012,Ramsey2016}, WaterMap \cite{Young2007-watermap,Abel2008-watermap}, SSTMap \cite{Haider2018-sstmap}, STOW \cite{Li2012-stow} and Solvaware \cite{Huggins2016-solvaware}.
The main limitation of those methods is that the entropy is approximated through an infinite expansion in terms of the density distribution. 
Due to the high computational effort required to converge the higher order terms, the entropy expansion is usually truncated after the first or, sometimes, the second \cite{Nguyen2016-gist-second-order,Waibl2022-gist-solvents} term, thereby introducing an error.
%TODO add MIE citation and context
Furthermore, IST does not consider the internal degrees of freedom of the solvent.
Therefore, IST-based approaches can only be used with sufficiently rigid solvents.
Nevertheless, they have been applied on a wide variety of problems involving both small molecules and biological systems as solute molecules.

GIST is an implementation of IST on a three-dimensional grid. Unlike the related method WaterMap \cite{Abel2008-watermap,Young2007-watermap}, it provides solvation thermodynamics on an entire grid, rather than only high-density clusters in a binding pocket.
This implies a higher versatility, but also puts more responsibility on the user in terms of correct post-processing.

Since the density is computed from an explicit solvent MD simulation, GIST provides an accurate picture of solvation at an atomistic level, even for demanding systems such as drug binding sites in proteins.
While the need for MD simulations comes at a large computational cost, GIST is still more efficient than end-state approaches such as TI \cite{Kirkwood1935-ti} or FEP \cite{Zwanzig1954-reweighting}, with applications to large systems such as serine proteases \cite{Kraml2019-gigist} or antibody fragments \cite{Waibl2021-gist-antibodies}.
On the other hand, the first-order approximation means that the solvation entropy is less accurately described than in TI or FEP \cite{Chen2021,Waibl2021-gist-salt}. 
Similar to the self-consistent integral theories, however, GIST only considers a single solute conformation and  multiple computations of representative structures are necessary to describe ensemble properties.

@@

We start with a basic GIST analysis of biotin.
Building on the basics, we present a more advanced GIST study of the streptavidin-biotin complex, reflecting on many common aspects of GIST studies.
We show how to interpret the three-dimensional contributions of solvation free energy in a binding pocket and around the ligand.
Additionally, we compute the binding contribution of the free energy of solvation, which requires accurate post-processing of the GIST outputs to avoid unfavorable summation of bias.
We believe that this example study covers a wide range of applications.

Furthermore, we discuss several technical aspects of GIST studies, such as the normalization of voxel values and how to avoid double counting of energy contributions. 
Additionally, we provide a Python library (\software{gisttools}) to unify the analysis of GIST data produced by various versions of GIST and make the post-processing of GIST results more accessible.

After completing this tutorial, we expect the reader to be able to run their own GIST study.

@@

\subsection{Background knowledge}
This tutorial is aimed at users with a solid knowledge of molecular dynamics (MD) simulations.

The user should be able to run MD simulations and analyze the resulting trajectories. 
The MD simulations presented in this tutorial are run with the Amber simulation engine and analyzed using AmberTools \software{cpptraj}. 
For a conceptual overview of running MD simulations with Amber, we recommend reading the Amber manual (\url{https://ambermd.org/Manuals.php}) \cite{amber24} and the Amber tutorials (\url{https://ambermd.org/Manuals.php}). 
Previous experience of GIST is not necessary, although we recommend reading the works referenced in section~\ref{sec:theory}.

@@

\subsection{Software/system requirements}
%Tutorials should clearly define what system and/or software requirements the researcher will need to complete the tutorial (e.g., VMD version 1.9 or newer, AMBER, etc.). Tutorials requiring specific software packages must provide instructions and files for the referenced version of the software.
The main implementation of GIST is part of the MD analysis software \software{cpptraj} released with AmberTools \cite{amber24, Case2023-ambertools}.

A recent version of \software{cpptraj} should be used. We recommend using at least version 6.15.2, which can be obtained from GitHub (\url{https://github.com/Amber-MD/cpptraj}).

@@

Furthermore, a recent Python version (3.6 or newer) should be available with the following packages installed:
\begin{itemize}
	\item \software{gisttools} (\url{https://github.com/liedllab/gisttools}, $\geq$ 0.2)
	\item \software{mdtraj} (\url{https://www.mdtraj.org/}, $\geq$ 1.9.7) \cite{McGibbon2015-mdtraj}
	\item \software{matplotlib} (\url{https://matplotlib.org/}, $\geq$ 3.7.0) \cite{Hunter2007-matplotlib}
	\item \software{numpy} (\url{https://numpy.org/}, $\geq$ 1.23.5) \cite{harris2020-numpy}
	\item \software{pandas} (\url{https://pandas.pydata.org/}, $\geq$ 1.5.3) \cite{pandas-2023_8239932}
\end{itemize}

If the reader prefers to skip the MD and GIST calculation, we also provide the GIST output files with this tutorial, such that the post-processing can be done without any expensive calculations.
We additionally provide a Jupyter notebook \cite{Kluyver2016-jupyter,Granger2021-jupyter} with the presented analyses. 
%We recommend JupyterLab to work with the notebook. Jupyter is available at \url{https://jupyter.org/}.

For the visualization of three-dimensional contributions we require molecular visualization software that can handle \software{OpenDX} files, such as \software{PyMol} \cite{pymol} or \software{VMD} \cite{vmd}.

Most of the visualisations and results in this tutorial can also be obtained using the \software{gistpp} program, which is available from \url{https://github.com/KurtzmanLab/Gist-Post-Processing} (code and documentation). \software{Gistpp} can directly manipulate \software{OpenDX} files from the GIST output. 
Its usage and some examples were shown in a previous tutorial on GIST \cite{Ramsey2016}.

@@

\begin{lstlisting}[style=cpptraj]
gist gridcntr <x> <y> <z> \
griddim <Nx> <Ny> <Nz> gridspacn <val>
\end{lstlisting}

Additionally, the GIST calculation should usually be preceded by preprocessing steps, such as centering the solute molecule in the unit cell or reimaging to account for periodic boundary conditions.
Often, the region of interest will be defined as the surrounding of a molecule, which might for example be a ligand in the binding pocket or the whole protein. 
Using \inlinecode{cpptraj's } \inlinecode{bounds} command, the region of interest can be output to a file. For example, here is the command for a region of \SI{10}{\angstrom} around the ligand LIG:

@@

\begin{lstlisting}[style=cpptraj]
gist gridcntr 10.5 20.1 30.1 griddim 30 30 30 \
	gridspacn 0.50 out gist.dat
\end{lstlisting}

This command tells \software{cpptraj} to run a GIST calculation using the specified region and print the output in a file called \inlinecode{gist.dat}.  An overview of how to work with \software{cpptraj} is provided in the program's documentation \cite{cpptraj_doc}, AMBER tutorials \cite{amber_tut_cpptraj} and on the AMBER-hub website \cite{amber_hub}.

@@

\subsubsection{Equilibration}
Equilibrate your solvated biotin in an NVT or NPT ensemble.
Even when running the simulation in NVT, it is advisable to include an NPT equilibration step to adjust the box size beforehand.
Use enough equilibration time to allow the solvent to relax. 
For water, \SI{0.5}{\nano\second} is usually sufficient to equilibrate the density, but narrow binding pockets can take longer.

A well-tested equilibration protocol can be found in the literature \cite{Roe2020-equilibration}.

\subsubsection{MD simulation}
\label{sec:md_simulation}
Now, you can run a restrained MD simulation based on the equilibrated structure of biotin for the subsequent GIST analysis.
Restraints should be used during the MD because IST theory is not defined for moving solute molecules.
We apply a harmonic restraint of \SI{100}{\kilo\calorie\per\mole\per\angstrom\squared} to all heavy atoms to keep the system in place.
We run \SI{100}{\nano\second} of simulation and store the conformations every \SI{100}{\pico\second}.
A simulation length of at least \SI{20}{\nano\second} is recommended to obtain statistically independent snapshots of the water solvent molecules.

An example Amber input file for GIST might look like this:
\begin{lstlisting}[style=amber-in]
restrained 100 ns NPT
&cntrl
	ntx=5, irest=1,
	ioutfm=1,
	ntb=2, iwrap=1,
	ntr=1, restraint_wt=100.0,
	restraintmask='!@H=&!:WAT',
	ntp=1, pres0=1.0, taup=1.0,
	ntc=2, ntf=2,
	ntt=3, tempi=300.0, temp0=300.0, gamma_ln=2,
	nstlim=50000000, dt=0.002,
	ntwr=50000, ntwx=5000, ntpr=5000,
/
\end{lstlisting}

This file will run a \SI{100}{\nano\second} NPT MD simulation while keeping all atoms except for hydrogens and waters restrained.
Note that these are very standard MD settings except for the restraints.

Run the MD using \software{pmemd.cuda}:
\begin{lstlisting}[style=bash]
pmemd.cuda \
-O \
-i 100ns-npt-restraint.in \
-o md-01.out \
-p solvated.parm7 \
-c EQUIL-DONE.rst \
-x md-01.nc \
-r md-01.ncrst \
-ref EQUIL-DONE.rst
\end{lstlisting}
\subsubsection{GIST analysis}

@@

\begin{lstlisting}[style=cpptraj]
parm solvated.parm7 
trajin md-01.nc 1 1 1 
autoimage !(:WAT) origin 
strip :WAT 
trajout biotin-centered.pdb 
run
\end{lstlisting}

In this code snippet, the first two lines load the topology and coordinates of the first frame of our MD simulation. 
In the third line, we reimage the coordinates to the first box and center the biotin at the origin. 
Then, we strip all water solvent molecules in the fourth line and write out a pdb file in the fith line. 
Finally, we start the analysis with the "run" command.

@@

Now, you can run the actual GIST analysis.
Reimage and center the solute molecule as above, to verify that its position matches the grid parameters you decided on.

We recommend using either the GPU implementation or MPI-accelerated PME implementation of GIST.
While the GPU version is faster when limited CPU resources are available, the PME implementation matches the energy of the Amber MD engine more closely.
The MPI implementation \cite{Roe2023-mpi-gist} scales well over multiple CPU cores and can be combined with either the GPU or PME implementations for maximum efficiency. 
A speed comparison of the various parallelization implementations is shown in \cite{Roe2023-mpi-gist}.

\begin{lstlisting}[style=cpptraj]
parm solvated.parm7
trajin md-01.nc
autoimage !(:WAT) origin
# optionally remove the 'pme' option
gist griddim 100 100 100 out gist.dat \
refdens 0.03287 pme
\end{lstlisting}
\subsubsection{Working with the GIST results}
Running the GIST analysis above generates an output data file (gist.dat) containing all calculated quantities for each voxel in the grid. 
Additionally as outlined in section \ref{sec:running_GIST}, various quantities are written to OpenDX files.

To interact with your GIST data, we recommend the gisttools library for python3. Open the GIST output with gisttools:
\begin{lstlisting}[style=python]
from gisttools.gist import load_gist_file
import matplotlib.pyplot as plt
gist = load_gist_file(
'gist.dat', struct='solute-centered.pdb',  
eww_ref=-9.539)
print(f'n_frames = {gist.n_frames}')
print(f'rho0 = {gist.rho0}')
\end{lstlisting}
All information from the output file is now accessible through the gist object. 
If the number of frames of the simulation (n\_frames) and the reference density (rho0) are not provided, they are automatically detected from the GIST data. 
We will discuss a way to similarly detect the reference energy of the solvent-solvent energy Eww in the advanced section.

The GIST data can be accessed and modified from the gist object in various ways.
Gisttools provides access to single or multiple voxels through a \code{.loc} property similiar to the pandas library. 
GIST quantity data for all voxels can be accessed by directly indexing the gist object, either by providing a single name or a list of names.
Indexing with \code{.loc} works by providing a voxel index (either a index number or list of numbers) and optionally a column index (column name or list of names). Indices start from 0 and indexing will return a single entry, pandas series or pandas dataframe depending on the query:
\begin{lstlisting}[style=python]
gist.loc[0, 'population'] # single value
gist.loc[0] # Series
gist.loc[[0,1], ['x', 'y', 'z']] # DataFrame
gist['dTSsix_norm'].head() # Series
gist[['Eww_norm', 'Esw_norm']].head() # DataFrame
\end{lstlisting}

Gisttools provides access to all quantities listed in the GIST output file.
Additionally, it calculates the total enthalpy per voxel (Eall) as a sum of Eww and Esw, and the free energy of solvation (A) by subtracting $\Delta$TSsix from Eall.

Information about the grid is stored in the grid property of the gist object, structural data is stored as an \software{mdtraj} trajectory in the struct property:

Gisttools also provides OpenDX writing capabilities to visualize the calculated or modified GIST data. For visualization, you'll most likely want to write out the density-normalized GIST quantities (X\_dens):

\begin{lstlisting}[style=python]
gist.save_dx('A_dens','A.dx')
\end{lstlisting}

Section~\ref{sec:visualization} discusses the visualization of OpenDX files.

@@

The choice of initial structure is crucial to the GIST results.
Crystal structures may not represent the dominant solution structure for a given target.
In our example, we only use a single equilibrated structure.
A more sophisticated approach would be to identify several dominant structures in the ensemble and run a separate MD simulation and GIST calculation for each structure. Such dominant structures could either be identified from experimental ensembles or by clustering an unrestrained MD simulation. 
When starting from an experimental structure, care has to be taken during the preparation step.
Make sure to choose a proper protonation state for all titratable moeties.
You will need to solvate your structure in a solvent box, which should be large enough to obtain unbiased long-range electrostatics and avoid solvent-mediated interactions of periodic images.
A good criterion for validating the box size is to check whether the solvent on the outside of the box behaves bulk-like according to the GIST output.
Note that, due to the restraints and finite sampling time, some solvent sites might be inaccessible to the bulk solvent.
To investigate such positions, make sure to explicitly place water solvent molecules there. 
For cases where the exact occupation of such sites is unknown, it may be helpful to include grand-canonical Monte Carlo (GCMC) or nonequilibrium candidate Monte Carlo (NCMC) sampling steps in the equilibration or production stage \cite{Ge-GCMC2022,Melling-NCMC2023}.

It is good practice to check your restrained MD simulation or solvent populations obtained by GIST to confirm whether all relevant sites were properly solvated.

@@

Here, we do not include any sampling of the conformational states.
It is therefore important that the molecular conformations in the complex and dissociated states match exactly.
Use the \software{equilibration.py} script (or your own procedure) to perform short NVT and NPT equilibration runs on the complex structure.
Note that the equilibration script contains unrestrained NPT steps, because restraining two molecules at the same time might interfere with the pressure equilibration.
Then, use \software{cpptraj} to produce biotin and streptavidin structures based on the complex by stripping the respective other molecule like this:
\begin{lstlisting}[style=cpptraj]
parm complex/solvated.parm7
trajin complex/equil/EQUIL-DONE.rst
strip ^2 parmout streptavidin/solvated.parm7
trajout streptavidin/pre-equil.ncrst
\end{lstlisting}
Adapt the above input file for biotin by striping molecule 1 instead of 2 and replacing the path names in the \inlinecode{strip} and  \inlinecode{trajout} commands.
Alternatively, we provide a small script named \software{cpptraj\_remove\_mol.sh} for this procedure.

After that, use \software{equilibration.py -R} to run short restrained 
equilibrations (where only the water is allowed to move) on the individual 
systems.
By using restraints, we can ensure that the monomer conformations match the complex.

@@

\begin{equation}
\label{eq_ewwref}
E_{ww}^\textit{ref} = \frac{\int{E_{ww}} \mathrm{d}x}{\left<N_\textit{solvent}\right>}
  = \frac{\sum_\textit{voxels}E_{ww}^\textit{dens} V_\textit{vox}}{\sum_\textit{voxels}{N_\textit{solvent}} / N_\textit{frames}}
\end{equation}

Alternatively, the solvent-solvent energy in a large GIST grid converges to the bulk value at sufficient distance to the solute molecule. 
Therefore, the reference energy can be obtained by binning the voxels by their distance to the solute molecule, and evaluate Equation~\ref{eq_ewwref} within each bin.
If this value converges with increasing distance from the solute, it can be used as $E_{ww}^\textit{ref}$.

@@

\begin{lstlisting}[style=python]
from gisttools.gist import load_gist_file
import matplotlib.pyplot as plt
gist = load_gist_file(
    'gist.dat', struct='solute-centered.pdb')
bins, eww = gist.rdf(
    'Eww_unref_norm', bins=100, rmax=20.,
    normalize='norm')
plt.plot(bins, eww)
eww_auto = gist.detect_reference_value()
plt.axhline(eww_auto, color='k')
plt.gca().set(
    xlabel='distance [A]', ylabel='E_ww per mol')
plt.show()
\end{lstlisting}

The expected output of this code is shown in Figure~\ref{fig_ewwref}.
%TODO redo figure caption, yaxis label
\begin{figure}
	\centering
	\includegraphics[width=0.8\linewidth]{figures/Eww_convergence.eps}
	\caption{Convergence of $E_{ww}$ in the complex calculation with increasing distance to the solute molecule (streptavidin-biotin). The horizontal line shows the automatically computed reference energy.}\label{fig_ewwref}
\end{figure}

After choosing an appropriate reference value, we need to subtract this value from $E_{ww}$.
Make sure to subtract the reference from the normalized (\_norm) data.
In \software{gisttools}, this is done simply by setting \inlinecode{gist.eww\_ref}.

Now, sum all free energy contributions to obtain the spatially resolved $\dgsolv$.
In \software{gisttools}, this is done automatically and can be accessed using the \inlinecode{A\_dens} and \inlinecode{A\_norm} data rows.

>>>>

On the other hand, methods based on IST generally use a molecular solvent distribution obtained from an MD simulation to compute the free energy of hydration from its enthalpic and entropic contributions, providing easier interpretability of the resulting distributions.
Examples include grid inhomogeneous solvation theory (GIST) \cite{Nguyen2012,Ramsey2016}, WaterMap \cite{Young2007-watermap,Abel2008-watermap}, SSTMap \cite{Haider2018-sstmap}, STOW \cite{Li2012-stow} and Solvaware \cite{Huggins2016-solvaware}.
The main limitation of those methods is that the entropy is approximated through an infinite expansion in terms of the density distribution. 
Within GIST and other more recent methods, these entropy contributions are calulated via a k-nearest-neighbor (kNN) approach.  
Due to the high computational effort required to converge the higher order terms, the entropy expansion is usually truncated after the first or, sometimes, the second \cite{Nguyen2016-gist-second-order,Waibl2022-gist-solvents} term, thereby introducing an error.
While there are approximations available to deal with those problems, most noteworthy the mutual information expansion (MIE) and Maximum Information Spanning Tree (MIST) approaches, they are not currently implemented in GIST. 
Recent kNN implementations to calculate solvation entropy further improve phase space sampling by relabeling the solvent molecules in the simulation, treating the solvent molecules as distinguishable \cite{reinhard2007estimation,heinz2019computing,heinz2021permut,fogolari2021optimal}.For a recent review of kNN methods to calculate entropy in a molecular context, see \cite{fogolari2024k}.\\
Furthermore, IST does not consider the internal degrees of freedom of the solvent.
Therefore, IST-based approaches can only be used with sufficiently rigid solvents.
Nevertheless, they have been applied on a wide variety of problems involving both small molecules and biological systems as solute molecules.\\
GIST is an implementation of IST on a three-dimensional grid. Unlike the related method WaterMap \cite{Abel2008-watermap,Young2007-watermap}, it provides solvation thermodynamics on an entire grid, rather than only high-density clusters in a binding pocket.
This implies a higher versatility, but also puts more responsibility on the user in terms of correct post-processing.
Since the density is computed from an explicit solvent MD simulation, GIST provides an accurate picture of solvation at an atomistic level, even for demanding systems such as drug binding sites in proteins.
While the need for MD simulations comes at a large computational cost, GIST is still more efficient than end-state approaches such as TI \cite{Kirkwood1935-ti} or FEP \cite{Zwanzig1954-reweighting}, with applications to large systems such as serine proteases \cite{Kraml2019-gigist} or antibody fragments \cite{Waibl2021-gist-antibodies}.
On the other hand, the first-order approximation means that the solvation entropy is less accurately described than in TI or FEP \cite{Chen2021,Waibl2021-gist-salt}. 
Similar to the self-consistent integral theories, however, GIST only considers a single solute conformation and  multiple computations of representative structures are necessary to describe ensemble properties.

@@

We start with a basic GIST analysis of biotin.
Building on the basics, we present a more advanced GIST study of the streptavidin-biotin complex, reflecting on many common aspects of GIST studies.
We show how to interpret the three-dimensional contributions of solvation free energy in a binding pocket and around the ligand.
Additionally, we compute the binding contribution of the free energy of solvation, which requires accurate post-processing of the GIST outputs to avoid unfavorable summation of bias.
We believe that this example study covers a wide range of applications.\\
Furthermore, we discuss several technical aspects of GIST studies, such as the normalization of voxel values and how to avoid double counting of energy contributions. 
Additionally, we provide a Python library (\software{gisttools}) to unify the analysis of GIST data produced by various versions of GIST and make the post-processing of GIST results more accessible.
After completing this tutorial, we expect the reader to be able to run their own GIST study.

@@

\subsection{Background knowledge}
This tutorial is aimed at users with a solid knowledge of molecular dynamics (MD) simulations.\\
The user should be able to run MD simulations and analyze the resulting trajectories. 
The MD simulations presented in this tutorial are run with the Amber simulation engine and analyzed using AmberTools \software{cpptraj}. 
For a conceptual overview of running MD simulations with Amber, we recommend reading the Amber manual (\url{https://ambermd.org/Manuals.php}) \cite{amber24} and the Amber tutorials (\url{https://ambermd.org/Manuals.php}). 
Previous experience of GIST is not necessary, although we recommend reading the works referenced in section~\ref{sec:theory}.

@@

\subsection{Software/system requirements}
%Tutorials should clearly define what system and/or software requirements the researcher will need to complete the tutorial (e.g., VMD version 1.9 or newer, AMBER, etc.). Tutorials requiring specific software packages must provide instructions and files for the referenced version of the software.
The main implementation of GIST is part of the MD analysis software \software{cpptraj} released with AmberTools \cite{amber24, Case2023-ambertools}.
A recent version of \software{cpptraj} should be used. We recommend using at least version 6.15.2, which can be obtained from GitHub (\url{https://github.com/Amber-MD/cpptraj}).

@@

Furthermore, a recent Python version (3.6 or newer) should be available with the following packages installed:
\begin{itemize}
	\item \software{gisttools} (\url{https://github.com/liedllab/gisttools}, $\geq$ 0.2)
	\item \software{mdtraj} (\url{https://www.mdtraj.org/}, $\geq$ 1.9.7) \cite{McGibbon2015-mdtraj}
	\item \software{matplotlib} (\url{https://matplotlib.org/}, $\geq$ 3.7.0) \cite{Hunter2007-matplotlib}
	\item \software{numpy} (\url{https://numpy.org/}, $\geq$ 1.23.5) \cite{harris2020-numpy}
	\item \software{pandas} (\url{https://pandas.pydata.org/}, $\geq$ 1.5.3) \cite{pandas-2023_8239932}
\end{itemize}
If the reader prefers to skip the MD and GIST calculation, we also provide the GIST output files with this tutorial, such that the post-processing can be done without any expensive calculations.
We additionally provide a Jupyter notebook \cite{Kluyver2016-jupyter,Granger2021-jupyter} with the presented analyses. 
%We recommend JupyterLab to work with the notebook. Jupyter is available at \url{https://jupyter.org/}.

For the visualization of three-dimensional contributions we require molecular visualization software that can handle \software{OpenDX} files, such as \software{PyMol} \cite{pymol} or \software{VMD} \cite{vmd}.
Most of the visualisations and results in this tutorial can also be obtained using the \software{gistpp} program, which is available from \url{https://github.com/KurtzmanLab/Gist-Post-Processing} (code and documentation). \software{Gistpp} can directly manipulate \software{OpenDX} files from the GIST output. 
Its usage and some examples were shown in a previous tutorial on GIST \cite{Ramsey2016}.

@@

\begin{lstlisting}[style=cpptraj]
gist gridcntr <x> <y> <z> \
griddim <Nx> <Ny> <Nz> gridspacn <val>
\end{lstlisting}
Additionally, the GIST calculation should usually be preceded by preprocessing steps, such as centering the solute molecule in the unit cell or reimaging to account for periodic boundary conditions.
Often, the region of interest will be defined as the surrounding of a molecule, which might for example be a ligand in the binding pocket or the whole protein. 
Using \inlinecode{cpptraj's } \inlinecode{bounds} command, the region of interest can be output to a file. For example, here is the command for a region of \SI{10}{\angstrom} around the ligand LIG:

@@

\begin{lstlisting}[style=cpptraj]
gist gridcntr 10.5 20.1 30.1 griddim 30 30 30 \
	gridspacn 0.50 out gist.dat
\end{lstlisting}
This command tells \software{cpptraj} to run a GIST calculation using the specified region and print the output in a file called \inlinecode{gist.dat}.  An overview of how to work with \software{cpptraj} is provided in the program's documentation \cite{cpptraj_doc}, AMBER tutorials \cite{amber_tut_cpptraj} and on the AMBER-hub website \cite{amber_hub}.

@@

\subsubsection{Equilibration}
Equilibrate your solvated biotin in an NVT or NPT ensemble.
Even when running the simulation in NVT, it is advisable to include an NPT equilibration step to adjust the box size beforehand.
Use enough equilibration time to allow the solvent to relax. 
For water, \SI{0.5}{\nano\second} is usually sufficient to equilibrate the density, but narrow binding pockets can take longer.
A well-tested equilibration protocol can be found in the literature \cite{Roe2020-equilibration}.

\subsubsection{MD simulation}
\label{sec:md_simulation}
Now, you can run a restrained MD simulation based on the equilibrated structure of biotin for the subsequent GIST analysis.
Restraints should be used during the MD because IST theory is not defined for moving solute molecules.
We apply a harmonic restraint of \SI{100}{\kilo\calorie\per\mole\per\angstrom\squared} to all heavy atoms to keep the system in place.
We run \SI{100}{\nano\second} of simulation and store the conformations every \SI{100}{\pico\second}.
A simulation length of at least \SI{20}{\nano\second} is recommended to obtain statistically independent snapshots of the water solvent molecules. 
An example Amber input file for GIST might look like this:
\begin{lstlisting}[style=amber-in]
restrained 100 ns NPT
&cntrl
	ntx=5, irest=1,
	ioutfm=1,
	ntb=2, iwrap=1,
	ntr=1, restraint_wt=100.0,
	restraintmask='!@H=&!:WAT',
	ntp=1, pres0=1.0, taup=1.0,
	ntc=2, ntf=2,
	ntt=3, tempi=300.0, temp0=300.0, gamma_ln=2,
	nstlim=50000000, dt=0.002,
	ntwr=50000, ntwx=5000, ntpr=5000,
/
\end{lstlisting}
This file will run a \SI{100}{\nano\second} NPT MD simulation while keeping all atoms except for hydrogens and waters restrained.
Note that these are very standard MD settings except for the restraints.
Run the MD using \software{pmemd.cuda}:
\begin{lstlisting}[style=bash]
pmemd.cuda \
-O \
-i 100ns-npt-restraint.in \
-o md-01.out \
-p solvated.parm7 \
-c EQUIL-DONE.rst \
-x md-01.nc \
-r md-01.ncrst \
-ref EQUIL-DONE.rst
\end{lstlisting}
\subsubsection{GIST analysis}

@@

\begin{lstlisting}[style=cpptraj]
parm solvated.parm7 
trajin md-01.nc 1 1 1 
autoimage !(:WAT) origin 
strip :WAT 
trajout biotin-centered.pdb 
run
\end{lstlisting}
In this code snippet, the first two lines load the topology and coordinates of the first frame of our MD simulation. 
In the third line, we reimage the coordinates to the first box and center the biotin at the origin. 
Then, we strip all water solvent molecules in the fourth line and write out a pdb file in the fith line. 
Finally, we start the analysis with the "run" command.

@@

Now, you can run the actual GIST analysis.
Reimage and center the solute molecule as above, to verify that its position matches the grid parameters you decided on.
We recommend using either the GPU implementation or MPI-accelerated PME implementation of GIST.
While the GPU version is faster when limited CPU resources are available, the PME implementation matches the energy of the Amber MD engine more closely.
The MPI implementation \cite{Roe2023-mpi-gist} scales well over multiple CPU cores and can be combined with either the GPU or PME implementations for maximum efficiency. 
A speed comparison of the various parallelization implementations is shown in \cite{Roe2023-mpi-gist}.

\begin{lstlisting}[style=cpptraj]
parm solvated.parm7
trajin md-01.nc
autoimage !(:WAT) origin
# optionally remove the 'pme' option
gist griddim 100 100 100 out gist.dat \
refdens 0.03287 pme
\end{lstlisting}
\subsubsection{Working with the GIST results}
Running the GIST analysis above generates an output data file (gist.dat) containing all calculated quantities for each voxel in the grid. 
Additionally as outlined in section \ref{sec:running_GIST}, various quantities are written to OpenDX files.
To interact with your GIST data, we recommend the gisttools library for python3. Open the GIST output with gisttools:
\begin{lstlisting}[style=python]
from gisttools.gist import load_gist_file
import matplotlib.pyplot as plt
gist = load_gist_file(
'gist.dat', struct='solute-centered.pdb',  
eww_ref=-9.539)
print(f'n_frames = {gist.n_frames}')
print(f'rho0 = {gist.rho0}')
\end{lstlisting}
All information from the output file is now accessible through the gist object. 
If the number of frames of the simulation (n\_frames) and the reference density (rho0) are not provided, they are automatically detected from the GIST data. 
We will discuss a way to similarly detect the reference energy of the solvent-solvent energy Eww in the advanced section.

The GIST data can be accessed and modified from the gist object in various ways.
Gisttools provides access to single or multiple voxels through a \code{.loc} property similiar to the pandas library. 
GIST quantity data for all voxels can be accessed by directly indexing the gist object, either by providing a single name or a list of names.
Indexing with \code{.loc} works by providing a voxel index (either a index number or list of numbers) and optionally a column index (column name or list of names). Indices start from 0 and indexing will return a single entry, pandas series or pandas dataframe depending on the query:
\begin{lstlisting}[style=python]
gist.loc[0, 'population'] # single value
gist.loc[0] # Series
gist.loc[[0,1], ['x', 'y', 'z']] # DataFrame
gist['dTSsix_norm'].head() # Series
gist[['Eww_norm', 'Esw_norm']].head() # DataFrame
\end{lstlisting}
Gisttools provides access to all quantities listed in the GIST output file.
Additionally, it calculates the total enthalpy per voxel (Eall) as a sum of Eww and Esw, and the free energy of solvation (A) by subtracting $\Delta$TSsix from Eall.

Information about the grid is stored in the grid property of the gist object, structural data is stored as an \software{mdtraj} trajectory in the struct property:
%TODO we are missing some code here
Gisttools also provides OpenDX writing capabilities to visualize the calculated or modified GIST data. For visualization, you'll most likely want to write out the density-normalized GIST quantities (X\_dens):

\begin{lstlisting}[style=python]
gist.save_dx('A_dens','A.dx')
\end{lstlisting}
Section~\ref{sec:visualization} discusses the visualization of OpenDX files.

@@

The choice of initial structure is crucial to the GIST results.
Crystal structures may not represent the dominant solution structure for a given target.
In our example, we only use a single equilibrated structure.
A more sophisticated approach would be to identify several dominant structures in the ensemble and run a separate MD simulation and GIST calculation for each structure. 
Such dominant structures could either be identified from experimental ensembles or by clustering an unrestrained MD simulation. 
When starting from an experimental structure, care has to be taken during the preparation step.
Make sure to choose a proper protonation state for all titratable moeties.
You will need to solvate your structure in a solvent box, which should be large enough to obtain unbiased long-range electrostatics and avoid solvent-mediated interactions of periodic images.
A good criterion for validating the box size is to check whether the solvent on the outside of the box behaves bulk-like according to the GIST output.
Note that, due to the restraints and finite sampling time, some solvent sites might be inaccessible to the bulk solvent.
To investigate such positions, make sure to explicitly place water solvent molecules there. 
For cases where the exact occupation of such sites is unknown, it may be helpful to include grand-canonical Monte Carlo (GCMC) or nonequilibrium candidate Monte Carlo (NCMC) sampling steps in the equilibration or production stage \cite{Ge-GCMC2022,Melling-NCMC2023}.
It is good practice to check your restrained MD simulation or solvent populations obtained by GIST to confirm whether all relevant sites were properly solvated.

@@

Here, we do not include any sampling of the conformational states.
It is therefore important that the molecular conformations in the complex and dissociated states match exactly.
Use the \software{equilibration.py} script (or your own procedure) to perform short NVT and NPT equilibration runs on the complex structure.
Note that the equilibration script contains unrestrained NPT steps, because restraining two molecules at the same time might interfere with the pressure equilibration.
Then, use \software{cpptraj} to produce biotin and streptavidin structures based on the complex by stripping the respective other molecule like this:
\begin{lstlisting}[style=cpptraj]
parm complex/solvated.parm7
trajin complex/equil/EQUIL-DONE.rst
strip ^2 parmout streptavidin/solvated.parm7
trajout streptavidin/pre-equil.ncrst
\end{lstlisting}
Adapt the above input file for biotin by striping molecule 1 instead of 2 and replacing the path names in the \inlinecode{strip} and  \inlinecode{trajout} commands.
Alternatively, we provide a small script named \software{cpptraj\_remove\_mol.sh} for this procedure.
After that, use \software{equilibration.py -R} to run short restrained 
equilibrations (where only the water is allowed to move) on the individual 
systems.
By using restraints, we can ensure that the monomer conformations match the complex.

@@

\begin{equation}
\label{eq_ewwref}
E_{ww}^\textit{ref} = \frac{\int{E_{ww}} \mathrm{d}x}{\left<N_\textit{solvent}\right>}
  = \frac{\sum_\textit{voxels}E_{ww}^\textit{dens} V_\textit{vox}}{\sum_\textit{voxels}{N_\textit{solvent}} / N_\textit{frames}}
\end{equation}
Alternatively, the solvent-solvent energy in a large GIST grid converges to the bulk value at sufficient distance to the solute molecule. 
Therefore, the reference energy can be obtained by binning the voxels by their distance to the solute molecule, and evaluate Equation~\ref{eq_ewwref} within each bin.
If this value converges with increasing distance from the solute, it can be used as $E_{ww}^\textit{ref}$.

@@

\begin{lstlisting}[style=python]
from gisttools.gist import load_gist_file
import matplotlib.pyplot as plt
gist = load_gist_file(
    'gist.dat', struct='solute-centered.pdb')
bins, eww = gist.rdf(
    'Eww_unref_norm', bins=100, rmax=20.,
    normalize='norm')
plt.plot(bins, eww)
eww_auto = gist.detect_reference_value()
plt.axhline(eww_auto, color='k')
plt.gca().set(
    xlabel='distance [A]', ylabel='E_ww per mol')
plt.show()
\end{lstlisting}
The expected output of this code is shown in Figure~\ref{fig_ewwref}.
%TODO redo figure caption, yaxis label
\begin{figure}
	\centering
	\includegraphics[width=0.8\linewidth]{figures/Eww_convergence.eps}
	\caption{Convergence of $E_{ww}$ in the complex calculation with increasing distance to the solute molecule (streptavidin-biotin). The horizontal line shows the automatically computed reference energy.}\label{fig_ewwref}
\end{figure}

After choosing an appropriate reference value, we need to subtract this value from $E_{ww}$.
Make sure to subtract the reference from the normalized (\_norm) data.
In \software{gisttools}, this is done simply by setting \inlinecode{gist.eww\_ref}.
Now, sum all free energy contributions to obtain the spatially resolved $\dgsolv$.
In \software{gisttools}, this is done automatically and can be accessed using the \inlinecode{A\_dens} and \inlinecode{A\_norm} data rows.
